Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-01 RevisedSubmission/hines_hetland_2017_gji.tex
did more

By choosing a FOGM process for $z_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
where we pick $\mitbf{\theta}$ to maximize the probability of drawing
the observed data, $\mitbf{d}_i^*$, from the random vector
$\mitbf{d}_i$. However, unlike the MLE method, the REML method
produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to estimate $\mitbf{\theta}$
at 38 continuous GNSS stations in the Pacific Northwest that are east
of 121$^\circ$W. These stations are sufficiently far from the
subduction zone that they are unlikely to contain transient signal
associated with SSEs, allowing us to ignore displacements from $u_i$
in $\mitbf{d}_i$. We assume the noise at these inland stations is
representative of the noise at all the stations considered in this
study, which is probably a poor assumption since the inland stations
are subject to distinctly different climatic conditions. Nonetheless,
we find $\mitbf{\theta}$ for each of the inland stations and for each
displacement component by maximizing the REML likelihood function
\begin{equation}\label{eq:REMLNoise}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*}
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}
In the above equation, $\data^*_i$ consists of the data for a single station.

@@

We use the REML method to determine the appropriate values for the
parameters in $X_i$ and $T_i$, $\mitbf{\theta} = \{\ell, \phi, \tau
\}$. We divide the GNSS data into seven subsets that are four months
long and each centered on the time of a SSE. The times of the seven
SSEs are determined with tremor records from \cite{Wech2010}. We
choose to make each data subsets four months long because it is long
enough to encompass a SSE in the Pacific Northwest, while it is still
computationally tractable to estimate $\mitbf{\theta}$. However, four
months is too short to resolve the sinusoids in $\data_i$, and they
are omitted from $\data_i$ in this REML analysis. We find the optimal
$\mitbf{\theta}$ for each subset of data, for each displacement
component, and for each choice of $T_i$. The REML likelihood function
that we are maximizing with respect to $\mitbf{\theta}$ is now
\begin{equation}\label{eq:REMLPrior} 
\mathcal{L}(\mitbf{\theta}) =
\left(\frac{\left| \G^T \G \right|}{(2\pi)^{n-6m}
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2} \data_i^{*T} \mitbf{K}(\mitbf{\theta}) \data_i^*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G
                            \left( \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right)^{-1}
                            \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1}
\end{equation}
and $\mitbf{\Sigma}(\mitbf{\theta}) = \mitbf{C}_{\eta_i} +
C_{u_i}(\points,\points;\mitbf{\theta})$. In the above equation,
$\data^*_i$ consists of a single subset of data. The estimated
parameters are summarized in Table 1. Based on the interquartile
ranges, the estimated parameters for the SE and Wendland covariance
functions do not vary significantly between SSEs. This suggests that
the median of the estimated parameters should be appropriate for all
the Pacific Northwest SSEs. For the IBM model, there are several
anomalously large values for $\ell$ and $\phi$, which is why the
interquartile ranges are large.

@@

A more intuitive way of deciding which function to use for $T_i$ is to
compare the posterior displacements and the observed displacements.
The posterior displacements consist of our estimate of transient
displacements, secular trends, and seasonal deformation. Specifically,
the posterior displacement vector is $\hat{\data}_i = \left(
u_i(\points) + \G\mitbf{m}_i \right) | \left( \data_i = \data_i^*
\right)$. Following a similar procedure as in Section
\ref{sec:Method}, it can be shown that $\hat{\data}_i$ has a
multivariate normal distribution with mean vector
\begin{equation}\label{eq:DataPredMean}
\mitbf{\mu}_{\hat{d}_i} = \left[\begin{array}{cc}
                          C_{u_i}(\points,\points) & \G \\
                          \end{array}\right]
                          \left[\begin{array}{cc}
                          \mitbf{C}_{\eta_i} + C_{u_i}(\points,\points) & \G \\
                          \G^T  & \zeros \\
                          \end{array}\right]^{-1}
                          \left[\begin{array}{c}
                          \data_i^* \\
                          \zeros \\
                          \end{array}\right]
\end{equation}  
and covariance matrix
\begin{equation}\label{eq:DataPredCov}
\mitbf{C}_{\hat{d}_i} = C_{u_i}(\points,\points) - 
                        \left[\begin{array}{cc}
                              C_{u_i}(\points,\points) & \G \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \mitbf{C}_{\eta_i} + C_{u_i}(\points,\points) & \G \\
                              \G^T  & \zeros \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_{u_i}(\points,\points) \\
                              \G^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\data}_i$ using the SE, Wendland, and IBM covariance
functions for $T_i$, and we use the median values from Table 1 for
$\mitbf{\theta}$. Figure \ref{fig:Fit} compares $\data_i^*$ to
$\hat{\data}_i$ for the winter 2015-2016 SSE at stations ALBH, P436,
and SC02. Based on Figure \ref{fig:Fit}, the chosen covariance
function for $T_i$ has almost no effect on $\hat{\data}_i$. The
posterior displacements for the IBM covariance function contain
slightly more high frequency, and perhaps spurious, features.
Regardless of the chosen covariance function, $\hat{\data}_i$ appears
to underestimate the rate of deformation during the SSE at stations
ALBH and SC02. The deformation at these two stations is particularly
rapid compared to the surrounding stations, and the misfit between
$\data_i^*$ and $\hat{\data}_i$ is likely due to over-smoothing. This
over-smoothing could indicate that the chosen values for $\tau$ or
$\ell$ are too large. However, $\hat{\data}_i$ does faithfully fit
$\data_i^*$ at the remaining stations, and so we do not attempt to
further refine the parameters.

@@

\begin{figure*}
\includegraphics{figures/signal_fit/signal-fit.pdf}
\caption{
Easting component of the observed displacements, $\data_\e^*$, and
posterior displacements, $\hat{\data}_\e$, during the winter 2015-2016
SSE. We show $\hat{\data}_\e$ for when $T_i$ is a SE, Wendland, and
IBM covariance function. The one standard deviation uncertainties are
shown for $\data_\e^*$ and the $\hat{\data}_\e$ that has a SE
covariance function for $T_i$. For clarity, uncertainties are not
shown for $\hat{\data}_\e$ when $T_i$ is an IBM or Wendland covariance
function, but the uncertainties are nearly equivalent to the
uncertainties for when $T_i$ is the SE covariance function. The
posterior displacements for the different covariance functions are all
practically indistinguishable.
}   
\label{fig:Fit}
\end{figure*}

>>>>

By choosing a FOGM process for $z_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
where we pick $\mitbf{\theta}$ to maximize the probability of drawing
the observed data, $\mitbf{d}_i^*$, from the random vector
$\mitbf{d}_i$. However, unlike the MLE method, the REML method
produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to estimate $\mitbf{\theta}$
at 38 continuous GNSS stations in the Pacific Northwest that are east
of 121$^\circ$W. These stations are sufficiently far from the
subduction zone that they are unlikely to record transient deformation
from SSEs, allowing us to ingore the term $u_i(\points)$ in
$\mitbf{d}_i$. We assume the noise at these inland stations is
representative of the noise at all the stations considered in this
study, which is probably a poor assumption since the inland stations
are subject to distinctly different climatic conditions. Nonetheless,
we find $\mitbf{\theta}$ for each of the inland stations and for each
displacement component by maximizing the REML likelihood function
\begin{equation}\label{eq:REMLNoise}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*}
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}
In the above equation, $\data^*_i$ consists of the data for a single station.

@@

We use the REML method to determine the appropriate values for the
parameters in $C_{u_i}$, which are $\ell$, $\phi$, and $\tau$. We
refer to these parameters collectively as $\mitbf{\theta}$. Just as in
Section \ref{sec:NoiseModel}, we pick $\mitbf{\theta}$ to maximize the
probability of drawing $\data_i^*$ from $\data_i$, but now we are
including $u_i(\points)$ in $\data_i$. and we are optimizing the
parameters for $C_{u_i}$ rather than $\mitbf{C}_{\eta_i}$. We divide
the GNSS data into seven subsets that are four months long and each
centered on the time of a SSE. The times of the seven SSEs are
determined with tremor records from \cite{Wech2010}. We find the
optimal $\mitbf{\theta}$ for each subset of data, for each
displacement component, and for each choice of $T_i$. The REML
likelihood function that we are maximizing with respect to
$\mitbf{\theta}$ is now
\begin{equation}\label{eq:REMLPrior} 
\mathcal{L}(\mitbf{\theta}) =
\left(\frac{\left| \G^T \G \right|}{(2\pi)^{n-6m}
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2} \data_i^{*T} \mitbf{K}(\mitbf{\theta}) \data_i^*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G
                            \left( \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right)^{-1}
                            \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1}
\end{equation}
and $\mitbf{\Sigma}(\mitbf{\theta}) = \mitbf{C}_{\eta_i} +
C_{u_i}(\points,\points;\mitbf{\theta})$. In the above equation,
$\data^*_i$ consists of a single subset of data. The estimated
parameters are summarized in Table 1. Based on the interquartile
ranges, the estimated parameters for the SE and Wendland covariance
functions do not vary significantly between SSEs. This suggests that
the median of the estimated parameters should be appropriate for all
the Pacific Northwest SSEs. For the IBM model, there are several
anomalously large values for $\ell$ and $\phi$, which is why the
interquartile ranges are large.

@@

A more intuitive way of deciding which function to use for $T_i$ is to
compare the posterior displacements and the observed displacements.
The posterior displacements consist of our estimate of transient
displacements, secular trends, and seasonal deformation. Specifically,
the posterior displacement vector is $\hat{\data}_i = \left(
u_i(\points) + \G\mitbf{m}_i \right) | \left( \data_i = \data_i^*
\right)$. Following a similar procedure as in Section
\ref{sec:Method}, it can be shown that $\hat{\data}_i$ has a
multivariate normal distribution with mean vector
\begin{equation}\label{eq:DataPredMean}
\mitbf{\mu}_{\hat{d}_i} = \left[\begin{array}{cc}
                          C_{u_i}(\points,\points) & \G \\
                          \end{array}\right]
                          \left[\begin{array}{cc}
                          \mitbf{C}_{\eta_i} + C_{u_i}(\points,\points) & \G \\
                          \G^T  & \zeros \\
                          \end{array}\right]^{-1}
                          \left[\begin{array}{c}
                          \data_i^* \\
                          \zeros \\
                          \end{array}\right]
\end{equation}  
and covariance matrix
\begin{equation}\label{eq:DataPredCov}
\mitbf{C}_{\hat{d}_i} = C_{u_i}(\points,\points) - 
                        \left[\begin{array}{cc}
                              C_{u_i}(\points,\points) & \G \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \mitbf{C}_{\eta_i} + C_{u_i}(\points,\points) & \G \\
                              \G^T  & \zeros \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_{u_i}(\points,\points) \\
                              \G^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\data}_i$ using the SE, Wendland, and IBM covariance
functions for $T_i$, and we use the median values from Table 1 for
$\mitbf{\theta}$. Figure \ref{fig:Fit} compares $\data_i^*$ to
$\hat{\data}_i$ during the winter 2015-2016 SSE at stations ALBH,
P436, and SC02, which are the stations that recorded the strongest
signal from this SSE. Based on Figure \ref{fig:Fit}, the chosen
covariance function for $T_i$ has almost no effect on $\hat{\data}_i$.
The posterior displacements for the IBM covariance function contain
slightly more high frequency, and perhaps spurious, features.
Regardless of the chosen covariance function, $\hat{\data}_i$ appears
to underestimate the rate of deformation during the SSE at stations
ALBH and SC02. The deformation at these two stations is particularly
rapid compared to the surrounding stations, and the misfit between
$\data_i^*$ and $\hat{\data}_i$ is likely due to over-smoothing. This
over-smoothing could indicate that the chosen values for $\tau$ or
$\ell$ are too large. However, $\hat{\data}_i$ does faithfully fit
$\data_i^*$ at the remaining stations, and so we do not attempt to
further refine the parameters.

@@

\begin{figure*}
\includegraphics{figures/signal_fit/signal-fit.pdf}
\caption{
Easting component of the observed displacements, $\data_\e^*$, and
posterior displacements, $\hat{\data}_\e$, during the winter 2015-2016
SSE. We show $\hat{\data}_\e$ for when $T_i$ is a SE, Wendland, and
IBM covariance function. The one standard deviation uncertainties are
shown for $\data_\e^*$ and the $\hat{\data}_\e$ that has a SE
covariance function for $T_i$. For clarity, uncertainties are not
shown for the $\hat{\data}_\e$ that have an IBM or Wendland covariance
function for $T_i$. The posterior displacements for the different
covariance functions are all practically indistinguishable.
}   
\label{fig:Fit}
\end{figure*}
